The non-random assembly of indirect interactions in  \nplant-pollinator networks

### 2022-07-20 Manuscript/Main_text/Main_text_Journal_of_Animal_Ecol.Rmd
edits

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (Table S2). This dataset consisted on 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; Table S3). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the _FD_ package version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the _maptree_ package version _1.4-7_ [@white2009].

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variablity and provided highly similar or identical networks to the original ones. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controlling for the variation at the study level.

@@

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = L - l_i / L - 1$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialisation (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2020] version 2.7.0.

@@

The hierarchical cluster analysis divided the dataset into five different groups with different and overlapping characteristics (Figs S4 and S5). The first group referred to as 'selfing herbs' consisted on herbs with hermaphroditic flowers with high levels of autonomous selfing. The second group named 'small outcrossing perennials' had small perennial species with a mix of life forms (i.e., trees shrubs and herbs) with outcrossing hermaphroditic flowers. The third group referred to as 'self-incompatible perennials with large flowers' comprised perennial species with a mixed of life forms and large self-incompatible hermaphroditic flowers with high number of ovules. The fourth group named 'tall plants with small unisexual flowers' had the tallest species, highest proportion of shrub and tree life forms, dioecious and monoecious breeding systems, small flowers and the highest numbers of flowers and inflorescences per plant. Finally, the fifth group named 'short-lived outcrossers with long zygomorphic flowers' consisted on small perennial and short-lived herbs with long self-compatible zygomorphic flowers that were unable to self-pollinate.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=8, fig.width=12, fig.cap="**Figure 5.** Z-score comparison (95% confidence level) of the observed and expected frequencies of motif combinations with four specific graphical examples of the most probable motifs of being observed for each category delimited by the z-score critical values: under-represented (red), no statistical difference (green) and over-represented (blue)."}

>>>>

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (Table S2). This dataset consisted of 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; Table S3). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the _FD_ package version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the _maptree_ package version _1.4-7_ [@white2009].

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variability and provided highly similar or identical networks to the original ones. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controlling for the variation at the study level.

@@

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialisation (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2020] version 2.7.0.

@@

The hierarchical cluster analysis divided the dataset into five different groups with different and overlapping characteristics (Figs S4 and S5). The first group referred to as 'selfing herbs' consisted of herbs with hermaphroditic flowers with high levels of autonomous selfing. The second group named 'small outcrossing perennials' had small perennial species with a mix of life forms (i.e., trees shrubs and herbs) with outcrossing hermaphroditic flowers. The third group referred to as 'self-incompatible perennials with large flowers' comprised perennial species with a mixed of life forms and large self-incompatible hermaphroditic flowers with high number of ovules. The fourth group named 'tall plants with small unisexual flowers' had the tallest species, highest proportion of shrub and tree life forms, dioecious and monoecious breeding systems, small flowers and the highest numbers of flowers and inflorescences per plant. Finally, the fifth group named 'short-lived outcrossers with long zygomorphic flowers' consisted of small perennial and short-lived herbs with long self-compatible zygomorphic flowers that were unable to self-pollinate.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=8, fig.width=12, fig.cap="**Figure 5.** Distribution of z-scores values after the statistical comparison of the observed probabilities of finding a subgraph combination in real networks and the expected probabilities (obtained by computer simulations). The resulting z-scores frequencies of this statiscal comparison are coloured based on a 95% confidence level: under-represented ($z_i^O < -1.96$, red), no statistical difference ($-1.96 \\geq z_i^O \\leq  1.96$, green) and over-represented ($z_i^O > 1.96$, blue). For instance, for a z-score value of 0 there were approximately 800 subgraph combinations that showed no statistical difference in the comparison between real and simulated networks. In addition, four specific graphical examples of the most probable subgraphs being observed on each category are also provided and the different plant and floral visitor groups are represented with distinct colour codes as indicated in the legend."}
